 

### 2020-12-21 vertical-thermal-review.Rmd
updated

The capacity of buffering largly depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions.
Increasing temperature has profound impacts on canopy cover such that temperate and tropical forest canopies are already or will soon be functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019]. Additionally, increased temperature is driving more frequent droughts causing hydraulic limitation in forests. Larger trees suffer more during drought, which may contribute to increasing tree mortality in forests around the world [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a; @mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al. 2015]. Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].

@@

Moderating capacity of under forest microclimate differs between forests with varying canopy structure, crown density, and seasons. Tropical, and temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures more than pine and boreal forests [De Frenne et al. 2019; @vonarxSpatiotemporalEffectsForest2012a]. This relationship is reveresed during the night, with pine forests maintaing warmer below canopy temperature than above [@vonarxSpatiotemporalEffectsForest2012a]

Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012] and coastal Brazil [@fauset_differences_2018, Tymen et al. 2017].

@@

Along with temperature, forest canopies moderate relative humidity in below-canopy microclimates. In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area. Moderating capacity was stronger under broadleaved and non-pine conifer forests than pine, and greater in wetter conditions, and in summer and autumn months than winter. Dry months showed less drastic minimum RH changes in below vs upper canopy  [@vonarxSpatiotemporalEffectsForest2012a].

@@

*Leaf area / characteristic dimension*:
- Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected (Leigh et al. 2017? https://onlinelibrary.wiley.com/doi/full/10.1111/pce.12857)
- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
Max transpiration rate - increases with canopy height (Kenzo et al. 2015).
Leaf width and characteristic dimension influence boundary layer resistance which generally is greater in wider understory leaves than upper canopy.
Additionally, increase in leaf area also influences leaf temperature by maintaining greater leaf temperature than leaves with lower leaf area {Wright et al. 2019}
-Boundary layer and thermal sensitivity also depends of leaf width, leaf size and structure, such as lobes, leaf elongation and serration, all dissapate heat more efficiently than non-lobed leaves, wide or non serrated leaves. Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies[Vogel 1970; Roth-Nebelsick 2001;
Schuepp 1993; Sack et al. 2006].

@@

**Many of the biophysical and trait variable reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019] (Fig. 3).
While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature.[@fauset_differences_2018]
Large leaves can be significantly cooler than $T_{air}$ under low radiation with stomata open, and significantly hotter under high radiation with stomata closed.  
Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm).
Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
However, counteracting this, lower wind speed in understory would reduce latent heat loss.
Thus, under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance; however, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.

@@

**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. **2?**).**
In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al. 1999).
Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).
**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-	(Slot et al. 2019) and refs therein
-	(Fauset et al. 2018)
Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016). There's some hint of this in Bolstad et al. 1999.

@@

*Stomatal conductance*

@@

**Photosynthesis is generally higher in sun leaves-- a fact that is well-established (Niinemets 2007 ; Kenzo et al. 2015 ; Slot et al. 2019; Chen et al. 2020) and observed in numerous field studies.**
This is primarily driven by the greater light available to sun leaves.
In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light,  including higher concentration of Rubsico, multiple layers of palisade parenchyma, and saturation level differs between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
The vertical gradient in photosythetic rates is also influenced by...
.[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N. Smith [agreed. Again the importance of differentiating between within-species and among species trait variation. Also; steepness of the vertical gradient in light and VPD. In very tall trees a gradient in water potential might cause a gradient in stomatal limitations of net photosynthesis]

@@

**Similar to photosynthesis, respiration tends to be higher in sun leaves (Chen et al. 2020), but its temperature sensitivity appears to be similar between between sun and shade leaves.**
(Bolstad et al. 1999) did a study at Coweeta, including elev gradient. 
Found higher respiration (at reference T) in canopy leaves (@bolstad_foliar_1999). (*Martijn has some unpublished data on this that we might include: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
At Coweeta, Q10 showed a variable pattern, with a slight tendency to increase down the canopy [@bolstad_foliar_1999]. (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
Combining reference respiration and $Q_{10}$, @bolstad_foliar_1999 found  a modest net decrease in R for understory leaves relative to canopy leaves as T increases (*see their Fig. 1*).
Thus, @bolstad_foliar_1999 found evidence of acclimation, but there remained a declining trend of respiration with elevation.

>>>>

The capacity of buffering largly depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions.
Increasing temperature has profound impacts on canopy cover such that temperate and tropical forest canopies are already or will soon be functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019]. Additionally, increased temperature is driving more frequent droughts causing hydraulic limitation in forests [ref]. Water limitation in forests alters species thermoregulation [@sastry_leaf_2018], also decreases tree canopy circumference [@aussenacInteractionsForestStands2000]. Evidence shows that larger trees suffer more than small trees during drought, which may contribute to increasing tree mortality in forests around the world [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a; @mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al. 2015]. Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020].

@@

Diurnal moderating capacity of under forest microclimate differs between forests with varying canopy structure, crown density, and seasons. Tropical, and temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures than pine and boreal forests. Increase in crown density and leaf area index increases decreases light and higher maximum temperature permeability in forests, thus maintaining lower maximum daytime temperature. Decrease in crown density and leaf area index such as in pine forests that have sparse stands decreases buffering capacity [De Frenne et al. 2019; @vonarxSpatiotemporalEffectsForest2012a]. This relationship is reveresed during the night among all forests, however, with pine forests maintaining warmer below canopy temperature than above[@vonarxSpatiotemporalEffectsForest2012a]

Similarly, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal Brazil [@fauset_differences_2018, Tymen et al. 2017].

@@

Along with temperature, forest canopies moderate relative humidity in below-canopy microclimates. In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area. Moderating capacity was stronger under broadleaved and non-pine conifer forests than pine, and greater in wetter conditions, and in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].

@@

*Leaf area / characteristic dimension*:
- Recent paper on influence on leaf size/shape in PCE – certain aspects of leaf shape were not as relevant as expected (Leigh et al. 2017? https://onlinelibrary.wiley.com/doi/full/10.1111/pce.12857)
- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
Max transpiration rate - increases with canopy height (Kenzo et al. 2015).
Leaf width and characteristic dimension influence boundary layer resistance which generally is greater in wider understory leaves than upper canopy.
Additionally, increase in leaf area also influences leaf temperature by maintaining greater leaf temperature than leaves with lower leaf area {Wright et al. 2019}
-Boundary layer resistance and thermal sensitivity also depends of leaf width, leaf size and structure. Increase in leaf width increases stomatal resistance-- dependent on stomtal activity and boundary layer resistance [@fauset_differences_2018] such as lobes, leaf elongation and serration, all dissapate heat more efficiently than non-lobed leaves, wide or non serrated leaves. Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies[Vogel 1970; Roth-Nebelsick 2001;
Schuepp 1993; Sack et al. 2006].

@@

**Many of the biophysical and trait variable reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019] (Fig. 3).
- As light availability typically increases with tree height, leaf temperature which is largely of function of radiation input, does as well [@michaletzPlantThermoregulationEnergetics2015]. Direct sunlight in upper canopy leaves increases leaf temperature, [sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress] more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature.[@fauset_differences_2018]
Large leaves can be significantly cooler than $T_{air}$ under low radiation with stomata open, and significantly hotter under high radiation with stomata closed.  
Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm).
Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
However, counteracting this, lower wind speed in understory would reduce latent heat loss.
Thus, under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance; however, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
- increase in air temperature, increases VPD, decrease in stomtal conductance during miday leading to decrease in CO2 in leaf = increase in leaf temperature in upper canopy. [@fauset_differences_2018]

@@

**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. **2?**).**
In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al. 1999).
Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).
**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil Forests despite much lower air temperature due to high solar radiation loads. However, leaf temperatures themselves were moderaed by air temperature. [@fauset_differences_2018].
-	(Slot et al. 2019) and refs therein
-	(Fauset et al. 2018)
Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016). There's some hint of this in Bolstad et al. 1999.

@@

*Stomatal conductance*
Photosynthesis can be largely constrained by stomatal conductance. 
-"Sun exposed outer canopy leaves in tropical trees can experience temperatures up 48 C during diurnal cycle, midday stomtal closure reduces transpitrational cooling" thus elevating leaf temperature greater than air temperature (Slot et al. 2019)

@@

**Photosynthesis is generally higher in sun leaves-- a fact that is well-established (Niinemets 2007 ; Kenzo et al. 2015 ; Slot et al. 2019; Chen et al. 2020) and observed in numerous field studies.**
This is primarily driven by the greater light available to sun leaves.
In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light,  including higher concentration of Rubsico, multiple layers of palisade parenchyma, and saturation level differs between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
The vertical gradient in photosythetic rates is also influenced by available concentration of N and P.
.[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N. Smith [agreed. Again the importance of differentiating between within-species and among species trait variation. Also; steepness of the vertical gradient in light and VPD. In very tall trees a gradient in water potential might cause a gradient in stomatal limitations of net photosynthesis]

@@

**Similar to photosynthesis, respiration tends to be higher in sun leaves (Chen et al. 2020), but its temperature sensitivity appears to be similar between between sun and shade leaves.**
At increasing temperatures, Photosynthetic System 2 [PS2] is particularly vulnerable. Damage to pS2  can lead to decreased electron transport rates and photosynthetic failure. Irreversible damage to PS2 is done at 40-60C [Baker 2008; @feeley_thermal_2020]
(Bolstad et al. 1999) did a study at Coweeta, including elev gradient. 
Found higher respiration (at reference T) in canopy leaves (@bolstad_foliar_1999). (*Martijn has some unpublished data on this that we might include: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
At Coweeta, Q10 showed a variable pattern, with a slight tendency to increase down the canopy [@bolstad_foliar_1999]. (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
Combining reference respiration and $Q_{10}$, @bolstad_foliar_1999 found  a modest net decrease in R for understory leaves relative to canopy leaves as T increases (*see their Fig. 1*)."Higher $Q_{10}$ values in the functional groups of upper canopy where temperatures are the highest may lead to exponential loses of C with increasing global temperatures depending upon the ability of canopy leaves to acclimate" [@cavaleri_foliar_2008]
Thus, @bolstad_foliar_1999 found evidence of acclimation, but there remained a declining trend of respiration with elevation.
